---
title: "Prediction Assignment "
author: "Hongyu Jin"
date: "January 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background.

Human Activity Recognition - HAR - has emerged as a key research area in the last years and is gaining increasing attention by the pervasive computing research community. Human activity recognition research has traditionally focused on discriminating between different activities, but they rarely quantify how well they do it. In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

This report aims to use machine learning algorithms to predict the class of exercise the individuals was performing by using measurements available from devices such as Jawbone Up, Nike Fuel Band, and Fitbit.



## Report.

### How the model was built.

Our outcome variable is classe, a factor variable with 5 levels. For this data set, "participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes." [1]
Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

### Cross-validation.

Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

### Expected out-of-sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

### Reasons for my choices

Our outcome variable "classe" is an unordered factor variable. Thus, we can choose our error type as 1-accuracy. We have a large sample size with N= 19622 in the Training data set. This allow us to divide our Training sample into subTraining and subTesting to allow cross-validation. Features with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables.
Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification [2]. Feature selection is inherent, so it is not so necessary at the data preparation phase. Thus, there won't be any feature selection section in this report.

### Code and Results.

Installing packages, loading libraries, and setting the seed for reproducibility:

```{r,knitr::opts_chunk$set(echo = TRUE)}
library(caret)
library(randomForest) #Random forest for classification and regression
library(rpart) # Regressive Partitioning and Regression trees
library(rpart.plot) # Decision Tree plot
library(ggplot2)
library(lattice)
# setting the overall seed for reproducibility.
set.seed(2658)
```
## Loading and Cleaning the Data sets.
First we want to load the data sets into R and make sure that missing values are coded correctly.
Irrelevant variables will be deleted.

```{r}
#Data loading

# Loading the training data set into R by replacing all missing values with "NA"
training <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!", ""))

# Loading the testing data set into R by replacing all missing values with "NA"
testing <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))

# Check dimensions for number of variables and number of observations
dim(training)
dim(testing)

# Delete columns with all missing values
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]

# Deleting the irrelevant variables : user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). 
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]

# new datasets:
dim(training)
dim(testing)

```

Following Code Results will be hidden from the report due to space considerations.
```{r,results='hide'}

head(training)
head(testing)

```

## Data Partitioning to allow cross-validation.
Training data was partitioned and preprocessed using the code described below.
The training data set contains 53 variables and 19622 obs.
The testing data set contains 53 variables and 20 obs.
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTest (25%).
This will be performed using random subsampling without replacement.

```{r}

subsamples <- createDataPartition(y=training$classe, p=0.75, list=FALSE)
subTraining <- training[subsamples, ] 
subTesting <- training[-subsamples, ]
dim(subTraining)
dim(subTesting)
```

Following Code Results will be hidden from the report due to space considerations.
```{r sub,results='hide'}
head(subTraining)
head(subTesting)
```

## Analysis of the Data.
 A, B, C, D and E are the five levels that contains in variable  "classe" . A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.

```{r}
plot(subTraining$classe, col="green", main="Plot of 5 levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency")
```



### Conclusion

From the above plot, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 4000 occurrences while level D is the least frequent with about 2500 occurrences.

## Prediction Models.

### First Prediction Model: Decision Tree.
```{r}
Firstmodel <- rpart(classe ~ ., data=subTraining, method="class")
# Prediction:
Predict1 <- predict(Firstmodel, subTesting, type = "class")

# Plotting the Decision Tree model.
rpart.plot(Firstmodel , main="Classification Tree",extra=102, under=TRUE, faclen=0)

```
```{r}
# Test results on  subTesting data set:
confusionMatrix(Predict1, subTesting$classe)

```

### Second prediction model: Random Forest.

```{r}
Secondmodel <- randomForest(classe ~. , data=subTraining, method="class")
# Prediction:
Predict2 <- predict(Secondmodel, subTesting, type = "class")
# Test results on subTesting data set:
confusionMatrix(Predict2, subTesting$classe)
```

## Conclusion.

As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.739 (95% CI: (0.727, 0.752)) for Decision Tree model. The random Forest model is chosen. The accuracy of the model is 0.995. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

## Final Submission.

Once, the predictions were obtained for the 20 test cases provided, the below shown script was used to obtain single text files to be uploaded to the courses web site to comply with the submission assigment.

# predict outcome levels on the original Testing data set using Random Forest algorithm

```{r}
Finalpredict <- predict(Secondmodel, testing, type="class")
Finalpredict

```

# Write files for submission

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(Finalpredict)
```

## Citation.

1. Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

2.  Krzysztof Grabczewski and Norbert Jankowski. Feature Selection with Decision Tree Criterion.

